Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}

### 2023-05-09 Papers/Paper CASC format/papercopy.tex
paper

\begin{document}
%
\title{Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}}
%
%\titlerunning{Abbreviated paper title}
%
\author{Diogo A. Gomes\inst{1}\orcidID{0000-0002-3129-3956} \and
	Friedemann Krannich\inst{1}\orcidID{1111-2222-3333-4444} \and
	Bashayer Majrashi\inst{1}\orcidID{0009-0008-3677-8064 } \and
	Ricardo de Lima Ribeiro\inst{1}\orcidID{0000-0002-2707-2210}}
%
\authorrunning{D.A. Gomes et al.}
%
\institute{King Abdullah University of Science and Technology (KAUST), AMCS/CEMSE Division, Thuwal 23955-6900, Saudi Arabia\
	\email{{diogo.gomes,friedemann.krannich,bashayer.majrashi, ricardo.ribeiro}@kaust.edu.sa}}
%
\maketitle
%
\begin{abstract}
	Many partial differential equations (PDEs) admit conserved quantities like mass or energy. These quantities are often essential for establishing well-posedness results. When approximating a PDE with a finite-difference scheme, it is crucial to determine whether related discretized quantities remain conserved over time by the scheme. Such conservation may ensure the stability of the numerical scheme. We present an algorithm for verifying the preservation of a polynomial quantity under a polynomial finite-difference scheme. Our schemes can be explicit or implicit, have higher-order time and space derivatives, and an arbitrary number of variables. Additionally, we introduce an algorithm for finding conserved quantities, given a scheme. We illustrate our algorithm with several finite-difference schemes. Our approach incorporates parametric G\"obner bases to handle parameters, ensuring accurate computation of conserved quantities.

	\keywords{Symbolic computations \and Finite-difference schemes \and Discrete variational derivative \and Discrete partial variational derivative \and Conserved quantities \and Implicit schemes \and Explicit schemes \and Parametric Groebner basis.}
\end{abstract}

\section{Introduction}
%\label{intro}
{\color{blue} add 
Comprehensive Gr{\"o}bner Systems 2222
}
In many partial differential equations (PDEs), conserved integral quantities play a crucial role, as they often represent properties of physical significance, such as energy or mass. For example, the advection equation preserves energy and the heat equation conserves mass. 
Since closed-form solutions for these PDEs are often unavailable, finite-difference schemes serve as a practical numerical alternative. Hence, it becomes imperative to determine whether the chosen scheme preserves these conserved quantities. This information is not only valuable for gauging the accuracy of the approximation, but also vital for evaluating its stability. Automating conservation computations in computer algebra systems streamlines the process, saves time, enhances accuracy, and may create new research opportunities.

@@

Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form
\begin{gather*}
	\frac{\partial }{\partial t}\Phi[u]+\div \Psi[u]=0
\end{gather*}
holding for all solutions $u$ of the considered PDE,
that induces the conserved quantity $\int \Phi[u]\ dx$ as
\begin{gather*}
	\frac{d}{dt}\int \Phi[u]\ dx=\int D_t\Phi[u]\ dx=-\int D_x\Psi[u]\ dx=0
\end{gather*}
assuming periodic boundary conditions in $x$.

@@

In this paper, we present an algorithm that verifies if a quantity is conserved over time under a given scheme. Our approach diverges from previous methods, as we do not attempt to construct discrete $\Phi$ and $\Psi$ using the scheme, but rather check if a given discrete $\Phi$ is conserved over time under the scheme. Typically, conserved quantities involve integral expressions or their discrete analogs, such as sums. Gomes et al. proposed algorithms for simplifying sums \cite{gomes20} and developed algorithms for detecting conserved quantities in PDEs and semi-discretized schemes \cite{gomes20}. We review these techniques in Section \ref{assp}.
To generalize these methods for cases without summing over all arguments of the involved functions, we introduce the discrete partial variational derivative (Section \ref{pvard}). Our main contribution is an algorithm for verifying the conservation of a quantity under a numerical scheme. Gerdt \cite{gerdt12} demonstrated that the quantity is conserved if its discrete time derivative belongs to the difference ideal generated by the scheme. However, some quantities may add only to a constant and be trivially preserved without being part of the difference ideal. Furthermore, Gerdt's algorithm may not always terminate, as the Gr{\"o}bner basis for the difference ideal might be infinite.
We address these issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) having a finite Gr{\"o}bner basis (Section \ref{algodetec}). Our algorithm is applicable to schemes that are explicit and implicit in time, can handle schemes with multiple higher-order time and space derivatives, multiple space dimensions, systems of equations, and schemes with parameters.
As the process of finding conserved quantities involves polynomial division, special care must be taken to ensure that the division is performed correctly with respect to the parameters. We employ parametric Gr{\"o}bner  bases, which are crucial for the proper treatment of parameters in our algorithm, see Section \ref{pgb}
 We have implemented this algorithm as a part of a {\sc Mathematica} package \cite{wolfram21}\footnote{The code is available upon request. For access, contact ricardo.ribeiro@kaust.edu.sa.}. Our examples include Burgers equation and a PDE system related to mean-field games (Section \ref{eaa}). In these examples, 
 we show that our code detects conserved quantities and can be used to select suitable schemes for time-implicit and time-explicit discretizations.

@@

	Let $N$, $d$ and $m$ be positive integers. The \emph{discrete torus} is the set $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space

@@

		\mathcal{F}(\mathcal{P}(\Ii,\Rr^m)):= \Bigg\{\mathcal{J}:\mathcal{P}(\Ii,\Rr^m)\to \Rr\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in\Ii}F_n[u],\\
		\text{ where }
		F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\\
		\text{ and }\{e^1,\ldots,e^k\}\subset\Zz^d,\ \text{G is smooth}
		\Bigg\}.

@@

@@

In the previous definition, $e^1,\ldots,e^k$ are not necessarily unit vectors and may vary between functionals.

@@

	\notag	=\sum_{n\in\Ii}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	=\sum_{n\in\Ii}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\

@@

	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))\Big)v(n)\\
	\notag	=:\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n).	

@@

The operator $\mathcal{V}(\mathcal{J})$ can be viewed as a representation of the derivative $D\mathcal{J}u$ with respect to the $L^2$ inner product $\langle w, v\rangle=\sum_{n\in \Ii} w_n v_n$, leading to the following definition.

@@

Because we assume all related functions are smooth, the discrete variational derivative always exists.

@@

		=\sum_{n\in\Ii}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\

@@

		=\sum_{n\in\Ii}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
		=\sum_{n\in\Ii}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)\\

@@

The algorithms for the simplification of sums presented in \cite{gomes20} relies on the following result, which 
provides a criterion to compare functionals and evaluate if they represent the same quantity.

\begin{theorem}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
	$\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.
	Conversely,
	$\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{theorem}
\begin{proof}
	We have
	\begin{gather*}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	\end{gather*}
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to determine whether different sums represent the same quantity.
\begin{example}
	Let $d=m=1$ and consider the functionals
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1)\text{ and }
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
	\end{gather*}
	It is clear that $\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity (by shifting $n$). We confirm this, using the discrete variational derivative:
	\begin{gather*}
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	\end{gather*}
	Hence, $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0$ and $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$. Therefore, both functionals represent the same quantity.
\end{example}

@@

Let $n=(n_1,\ldots,n_d)$ be the variables for the functions in $\mathcal{P}(\Ii,\Rr^m)$. We call $n$ the space variables. Let $l\in\Nn$ and call $t:=(t_1,\ldots,t_l)\in\Nn^l$ the time variables. For generality, we consider the vectorial time case, while later in this paper, we focus specifically on the case when $l=1$.
The analog to Definition \ref{d1} is the following. 
\begin{definition}
	Let $\mathcal{P}(\Ii\times\Nn^l,\Rr^m):=\{u\ |\ u(\cdot,t)\in\mathcal{P}(\Ii,\Rr^m)\ \forall t\in\Nn^l \}$.
	Functions in $\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$ extend to $\Zz^d\times\Nn^l$ through  periodicity in the space variables.
	Let $\mathcal{D}(\Nn^l):=\{u\ |\ u:\Nn^l\rightarrow\Rr\}$.
\end{definition}
We can regard $\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$ as the set of functions depending in space and time and $\mathcal{D}(\Nn^l)$ as the functions depending only on time.

Next, we examine functions that involve summation only over spatial variables, extending the concept presented in Definition \ref{d2}.
\begin{definition}
	\label{quantityfunc2}
	The space of functions $\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$ is
	\begin{gather*}
		\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l)):=\\	
		\Bigg\{\mathcal{J}:\mathcal{P}(\Ii\times\Nn^l)\rightarrow\mathcal{D}(\Nn^l) \Bigg|\ \mathcal{J}[u](t)=\sum_{n\in\Ii}F_{(n,t)}[u],\\
		\text{where } F_{(n,t)}[u]=G(u((n,t)+e^1),u((n,t)+e^2),\ldots,u((n,t)+e^k)),\\
		\text{and }
		\{e^1,\ldots,e^k\}\subset\Zz^{d+l},\text{G is a polynomial in k variables}\Bigg\}.
	\end{gather*}
	Here, $(n,t):=(n_1,\ldots,n_d,t_1,\ldots,t_l)\in\Ii\times\Nn^l$.
\end{definition}

@@

	+D_2G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^2)\\+\ldots

@@

	=\sum_{\tilde{n}}D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1,\ldots,t_{l})}(s_1-e^1_{d+1},\ldots,s_l-e^1_{d+l})v(\tilde{n})\\
	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1,\ldots,t_{l})}(s_1-e^2_{d+1},\ldots,s_l-e^2_{d+l})v(\tilde{n})\\+\ldots
	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1,\ldots,t_{l})}(s_1-e^k_{d+1},\ldots,s_l-e^k_{d+l})v(\tilde{n})\\

@@

	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1+e^2_{d+1},\ldots,t_{l}+e^2_{d+l})}(s_1,\ldots,s_l)\\+\ldots

@@

		D^1(\mathcal{J})[u](v)(t)\\
		=\frac{d}{d\epsilon} \sum_{n,s}((u(n,s+1)+\epsilon v(n,s+1))-(u(n,s)+\epsilon v(n,s)))\delta_t(s)\\
		=\sum_{n,s}v(n,s+1)\delta_t(s)-v(n,s)\delta_t(s)=\sum_{n,s}v(n,s)\delta_t(s-1)-v(n,s)\delta_t(s)\\
		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s)=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(n,s,t)v(n,s).

@@

We have a result similar to Theorem \ref{theo1} for the discrete partial variational derivative:

@@

\section{Difference schemes and algebra}
\label{algodetec}
In this section, we formally define numerical schemes and introduce the necessary tools from difference algebra for our algorithm.
Throughout this paper, we assume that there is only one time variable $t$ ($l=1$).
\begin{definition}
	The \emph{space of schemes} is
	\begin{gather*}
		\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr)):=\\
		\Bigg\{H:\mathcal{P}(\Ii\times\Nn,\Rr^m)\rightarrow\mathcal{P}(\Ii\times\Nn,\Rr)\ \Bigg|\\
		H[u](n,t)=H(u(n+\tilde{e}^1,t+\tilde{l}^1),\ldots,u(n+\tilde{e}^r,t+\tilde{l}^r)),\\
		\text{where }
		\{\tilde{e}^1,\ldots,\tilde{e}^r\}\subset\Zz^d,\ \{\tilde{l}^1,\ldots,\tilde{l}^r\}\subset\Zz,\\
		\text{H is a polynomial in r variables}
		\Bigg\}.
	\end{gather*}
	A \emph{scheme} is a set of functions
	\begin{gather*}
		\{H^1,\ldots,H^i\}\subset\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr))
	\end{gather*}
	that represent the equations $\{H^1[u]=0,\ldots,H^i[u]=0\}$ for $u\in\mathcal{P}(\Ii\times\Nn,\Rr^m)$, holding pointwise for all points in $\Ii\times\Nn$.
\end{definition}
If a scheme contains a single function $H$, we call $H$ the scheme. Sometimes, we also call the expression $H[u]$ the scheme.
\begin{remark}
	In our examples, we only consider finite-difference schemes with fixed step sizes $\Delta x=\Delta t=1$. However, our algorithms and our code handle parameters. Hence, our results can be generalized to schemes with other step sizes.
\end{remark}
\begin{example}
	\label{ee}
Consider the heat equation	$u_t-u_{xx}=0$ \cite{smoller94}.  A possible scheme is
\[	
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
\]
A solution of the corresponding numerical approximation is a function $u\in \mathcal{P}(\Ii\times\Nn,\Rr),$ that satisfies
\[
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t))=0
\]
for all $n\in \Ii$ and $t\in \Nn$. 
\end{example}

>>>>

\begin{document}
%
\title{Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}}
%
%\titlerunning{Abbreviated paper title}
%
\author{Diogo A. Gomes\inst{1}\orcidID{0000-0002-3129-3956} \and
	Friedemann Krannich\inst{1}\orcidID{0000-0002-8642-4319} \and
	Bashayer Majrashi\inst{1}\orcidID{0009-0008-3677-8064 } \and
	Ricardo de Lima Ribeiro\inst{1}\orcidID{0000-0002-2707-2210}}
%
\authorrunning{D.A. Gomes et al.}
%
\institute{King Abdullah University of Science and Technology (KAUST), AMCS/CEMSE Division, Thuwal 23955-6900, Saudi Arabia\
	\email{{diogo.gomes,friedemann.krannich,bashayer.majrashi, ricardo.ribeiro}@kaust.edu.sa}}
%
\maketitle
%
\begin{abstract}
	Many partial differential equations (PDEs) admit conserved quantities like mass or energy. These are often essential for establishing well-posedness results. When approximating a PDE with a finite-difference scheme, it is crucial to determine whether related discretized quantities remain conserved by the scheme. Such conservation may ensure the stability of the numerical scheme. We present an algorithm for verifying the preservation of a polynomial quantity under a polynomial finite-difference scheme. Our schemes can be explicit or implicit, have higher-order time and space derivatives, and  arbitrary number of variables. Additionally, we introduce an algorithm for finding conserved quantities. We illustrate our algorithm in several finite-difference schemes. Our approach incorporates parametric G\"obner bases to handle parameters, ensuring accurate computation of conserved quantities.

	\keywords{Finite-difference schemes \and Discrete variational derivative \and Discrete partial variational derivative \and Conserved quantities \and Implicit schemes \and Explicit schemes \and Parametric Groebner systems.}
\end{abstract}

\section{Introduction}
%\label{intro}
{\color{blue} add 
Comprehensive Gr{\"o}bner Systems 2222
}
In many partial differential equations (PDEs), conserved integral quantities play a crucial role, as they often represent properties of physical significance, such as energy or mass. For example, the advection equation preserves energy and the heat equation conserves mass. 
Since closed-form solutions for these PDEs are often unavailable, finite-difference schemes serve as a practical numerical alternative. Hence, it becomes imperative to determine whether the chosen scheme preserves these conserved quantities. This information is not only valuable for gauging the accuracy of the approximation, but also vital for evaluating its stability. Automating conservation computations in computer algebra systems streamlines the process, saves time and enhances accuracy.

@@

Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form
$
	\frac{\partial }{\partial t}\Phi[u]+\div \Psi[u]=0
$
holding for all solutions $u$ of the considered PDE,
that induces the conserved quantity $\int \Phi[u]\ dx$ as
$
	\frac{d}{dt}\int \Phi[u]\ dx=\int D_t\Phi[u]\ dx=-\int D_x\Psi[u]\ dx=0
$
assuming periodic boundary conditions in $x$.

@@

In this paper, we present an algorithm that verifies if a quantity is conserved over time under a given scheme. Our approach diverges from previous methods, as we do not attempt to construct discrete $\Phi$ and $\Psi$ using the scheme, but rather check if a given discrete $\Phi$ is conserved over time under the scheme. Typically, conserved quantities involve integral expressions or their discrete analogs, such as sums. Gomes et al. proposed algorithms for simplifying sums \cite{gomes20} and developed algorithms for detecting conserved quantities in PDEs and semi-discretized schemes \cite{gomes20}. We review these techniques in Section \ref{assp}.
To generalize these methods for cases without summing over all arguments of the involved functions, we introduce the discrete partial variational derivative (Section \ref{pvard}). Our main contribution is an algorithm for verifying the conservation of a quantity under a numerical scheme. Gerdt \cite{gerdt12} demonstrated that the quantity is conserved if its discrete time derivative belongs to the difference ideal generated by the scheme. However, some quantities may add only to a constant and be trivially preserved without being part of the difference ideal. Furthermore, Gerdt's algorithm may not always terminate, as the Gr{\"o}bner basis for the difference ideal might be infinite.
We address these issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) having a finite Gr{\"o}bner basis (Section \ref{algodetec}). Our algorithm is applicable to schemes that are explicit and implicit in time, can handle schemes with multiple higher-order time and space derivatives, multiple space dimensions, systems of equations, and schemes with parameters.
As the process of finding conserved quantities involves polynomial division, special care must be taken to ensure that the division is performed correctly with respect to the parameters. We employ parametric Gr{\"o}bner  bases, which are crucial for the proper treatment of parameters in our algorithm, see Section \ref{pgs}.
 We have implemented this algorithm as a part of a {\sc Mathematica} package \cite{wolfram21}\footnote{The code is available upon request. For access, contact ricardo.ribeiro@kaust.edu.sa.}. Our examples include Burgers equation and a PDE system related to mean-field games (Section \ref{eaa}). In these examples, 
 we show that our code detects conserved quantities and can be used to select suitable schemes for time-implicit and time-explicit discretizations.

@@

	Let $N$, $d$ and $m$ be positive integers. The \emph{discrete torus} is the set $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Let

@@

		\mathcal{F}(\mathcal{P}(\Ii,\Rr^m)):= \Bigg\{\mathcal{J}:\mathcal{P}(\Ii,\Rr^m)\to \Rr\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in\Ii}F_n[u],\text{ where for} G \ \text{ smooth} \\
		F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))
		\text{ and }\{e^1,\ldots,e^k\}\subset\Zz^d
		\Bigg\}, 

@@

where $e^1,\ldots,e^k$ are may not be unit vectors and may vary between functionals.

@@

@@

%	\notag	=\sum_{n\in\Ii}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
%	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	%=\sum_{n\in\Ii}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
%	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	%\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\

@@

%	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	\ldots+D_kG(u(n+e^1-e^k),\ldots,u(n))\Big)v(n)
=:\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n).	

@@

The operator $\mathcal{V}(\mathcal{J})$ represents the derivative $D\mathcal{J}u$ with respect to the $L^2$ inner product $\langle w, v\rangle=\sum_{n\in \Ii} w_n v_n$, leading to the following definition.

@@

Because here all functions are smooth, the discrete variational derivative exists.

@@

	%	=\sum_{n\in\Ii}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\

@@

	%	=\sum_{n\in\Ii}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
	%	=\sum_{n\in\Ii}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
	%	=\sum_{n\in\Ii}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)

@@

The algorithms for simplification of sums in \cite{gomes20} rely on the following result, which 
gives a criterion to decide if two functionals represent the same quantity.

\begin{theorem}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
	$\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.
	Conversely,
	$\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{theorem}
\begin{proof}
	We have
	\begin{gather*}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]=
		\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	\end{gather*}
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to determine whether different sums represent the same quantity.
\begin{example}
	Let $d=m=1$ and consider the functionals
$
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1)$ and $
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
$
By shifting $n$, we see that	$\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity . Accordingly, $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$, and the discrete variational derivative must vanish
	\begin{gather*}
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	\end{gather*}
%	Hence, $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0$
%	and $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$. Thus, the functionals agree.
	%	both functionals represent the same quantity.
\end{example}

@@

Let $n=(n_1,\ldots,n_d)$, the space variables, be the variables for the functions in $\mathcal{P}(\Ii,\Rr^m)$. Let $l\in\Nn$ and call $t:=(t_1,\ldots,t_l)\in\Nn^l$ the time variables. For generality, we consider the vectorial time case, while later in this paper, we focus on the case when $l=1$.
The analog to Definition \ref{d1} is the following. 
\begin{definition}
	Let $\mathcal{P}(\Ii\times\Nn^l,\Rr^m):=\{u\ |\ u(\cdot,t)\in\mathcal{P}(\Ii,\Rr^m)\ \forall t\in\Nn^l \}$.
	Functions in $\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$ extend to $\Zz^d\times\Nn^l$ through  periodicity in the space variables.
	Let $\mathcal{D}(\Nn^l):=\{u\ |\ u:\Nn^l\rightarrow\Rr\}$.
\end{definition}
We can regard $\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$ as the set of functions depending in space and time and $\mathcal{D}(\Nn^l)$ as the functions depending only on time.

Next, we examine functions that involve summation only over spatial variables, extending the concept presented in Definition \ref{d2}.
\begin{definition}
	\label{quantityfunc2}
	The space of functions $\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$ is the set
	\begin{gather*}
	%	\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l)):=\\	
		\Bigg\{\mathcal{J}:\mathcal{P}(\Ii\times\Nn^l)\rightarrow\mathcal{D}(\Nn^l) \Bigg|\ \mathcal{J}[u](t)=\sum_{n\in\Ii}F_{(n,t)}[u],\\
		\text{where } F_{(n,t)}[u]=G(u((n,t)+e^1),u((n,t)+e^2),\ldots,u((n,t)+e^k)),\\
		\text{and }
		\{e^1,\ldots,e^k\}\subset\Zz^{d+l},\text{G is a polynomial in k variables}\Bigg\}.
	\end{gather*}
	Here, $(n,t):=(n_1,\ldots,n_d,t_1,\ldots,t_l)\in\Ii\times\Nn^l$.
\end{definition}

@@

%	+D_2G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^2)\\
+\ldots

@@

%	=\sum_{\tilde{n}}D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1,\ldots,t_{l})}(s_1-e^1_{d+1},\ldots,s_l-e^1_{d+l})v(\tilde{n})\\
%	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1,\ldots,t_{l})}(s_1-e^2_{d+1},\ldots,s_l-e^2_{d+l})v(\tilde{n})\\+\ldots
%	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1,\ldots,t_{l})}(s_1-e^k_{d+1},\ldots,s_l-e^k_{d+l})v(\tilde{n})\\

@@

%	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1+e^2_{d+1},\ldots,t_{l}+e^2_{d+l})}(s_1,\ldots,s_l)\\+\ldots

@@

		D^1(\mathcal{J})[u](v)(t)
		=\frac{d}{d\epsilon} \sum_{n,s}((u(n,s+1)+\epsilon v(n,s+1))-(u(n,s)+\epsilon v(n,s)))\delta_t(s)=\\
		\sum_{n,s}v(n,s+1)\delta_t(s)-v(n,s)\delta_t(s)%=\sum_{n,s}v(n,s)\delta_t(s-1)-v(n,s)\delta_t(s)\\
		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s).%=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(n,s,t)v(n,s).

@@

The analog to Theorem \ref{theo1} for the discrete partial variational derivative is:

@@

\section{Difference schemes and algebra}
\label{algodetec}
In this section, we formally define numerical schemes and introduce the tools from difference algebra for our algorithm.
Throughout this paper, we assume that there is only one time variable $t$ ($l=1$).
\begin{definition}
	The \emph{space of schemes}, $	\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr))$ is the set
	\begin{gather*}
		\Bigg\{H:\mathcal{P}(\Ii\times\Nn,\Rr^m)\rightarrow\mathcal{P}(\Ii\times\Nn,\Rr)\ \Bigg|\\
		H[u](n,t)=H(u(n+\tilde{e}^1,t+\tilde{l}^1),\ldots,u(n+\tilde{e}^r,t+\tilde{l}^r)),\\
		\text{where }
		\{\tilde{e}^1,\ldots,\tilde{e}^r\}\subset\Zz^d,\ \{\tilde{l}^1,\ldots,\tilde{l}^r\}\subset\Zz,
		\text{H is polynomial}
		\Bigg\}.
	\end{gather*}
	A \emph{scheme} is a set of functions
$
		\{H^1,\ldots,H^i\}\subset\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr))
$
	that represent the equations $\{H^1[u]=0,\ldots,H^i[u]=0\}$ for $u\in\mathcal{P}(\Ii\times\Nn,\Rr^m)$, holding pointwise for all points in $\Ii\times\Nn$.
\end{definition}
If a scheme contains a single function $H$, we call $H$ the scheme. Sometimes, we also call the expression $H[u]$ the scheme.
%\begin{remark}
	In our examples, we only consider finite-difference schemes with fixed step sizes $\Delta x=\Delta t=1$. However, our algorithms and our code handle parameters. Hence, our results can be generalized to schemes with other step sizes.
%\end{remark}
\begin{example}
	\label{ee}
Consider the heat equation	$u_t-u_{xx}=0$ \cite{smoller94}.  A possible scheme is
$
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
$
A solution of the corresponding numerical approximation is a function $u\in \mathcal{P}(\Ii\times\Nn,\Rr),$ that satisfies
$
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t))=0
$
for all $n\in \Ii$ and $t\in \Nn$. 
\end{example}
